---
title: "Pipeline_Mental-disorders-STRs"
author: "Sergio Pérez Oliveira"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 3
  pdf_document:
    toc: true
    number_sections: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploring the role of CAG repeats in *HTT*, *ATXN1* and *ATXN2* genes in the genetic architecture of mental disorders: schizophrenia and bipolar disorder.

The project aims to evaluate the role of intermediate-length **CAG repeats** in *HTT*, *ATXN1 (SCA1)*, and *ATXN2 (SCA2)* genes in the etiology of **schizophrenia (SCZ)** and **bipolar disorder (BD)**. We perform association analyses, regression models, and effect size calculations using R to investigate potential modifying roles of these STR loci in mental illness phenotypes.

## 01_Environment.R

Load and preprocess datasets for analysis of frequencies and correlations in mental disorder cohorts.

```{r}
required_packages <- c(
  "pwr", "rcompanion","writexl", "FSA", "installr", "xlsx", "dplyr", "readxl", "ggplot2",
  "corrplot", "devtools", "ggpubr", "lsr", "Rcmdr", "survival", "KMsurv",
  "survMisc", "survminer", "ggfortify", "flexsurv", "actuar", "nortest",
  "rstatix", "gtsummary", "car", "DataExplorer", "effectsize", "lmtest", 
  "nnet","igraph","tidyverse","ggraph","tidygraph","visNetwork"
)

install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg)
    }
    library(pkg, character.only = TRUE)
  }
}

install_and_load(required_packages)

# Input section ----
# Select working directory interactively (select the main branch as directory)
wd<-choose.dir()
setwd<-wd
cat("Please select Mental disorders dataset (Excel):\n")
MENTAL <- read_excel(file.choose()) |> 
  select(-1) |> 
  select(-ADMISSION_AGE)

cat("Please select Control dataset (Excel):\n")
CONTROLS <- read_excel(file.choose())[-1, -1] |> 
  rename(AGE = DEATH_AGE)

# Functions ----

# Function to calculate mean and SD by category
mean_sd <- function(cohort, category, numeric_var, result_name) {
  categories <- unique(cohort[[category]])
  results <- data.frame()
  for (c in categories) {
    values <- na.omit(cohort[[numeric_var]][cohort[[category]] == c])
    results <- rbind(results, data.frame(
      CATEGORY = c,
      MEAN = round(mean(values), 2),
      SD = round(sd(values), 2)
    ))
  }
  colnames(results) <- c(category, paste(result_name, "MEAN"), paste(result_name, "SD"))
  return(results)
}

# Function to perform chi-square or Fisher's exact test with pairwise comparisons
my_function <- function(matrix) {
  chi_test <- chisq.test(matrix, correct = FALSE)
  perc_expected <- sum(chi_test$expected < 5) / length(chi_test$expected)
  
  if (perc_expected > 0.2) {
    fisher_test <- fisher.test(matrix)
    print("Fisher's test results:")
    print(fisher_test)
    test_result <- fisher_test$p.value
  } else {
    print("Chi-square test results:")
    print(chi_test)
    test_result <- chi_test$p.value
  }
  
  cat("\nMultiple comparisons:\n")
  print(pairwiseNominalIndependence(matrix, fisher = TRUE, chisq = TRUE, method = "holm"))
}

# Frequency table printing function
TABLE <- function(data, var1, var2, tableName) {
  cat(paste0("\nFrequency table: ", tableName, "\n"))
  table_mat <- xtabs(as.formula(paste("~", var1, "+", var2)), data = data)
  table_mat <- as.matrix(table_mat)
  table_mat[is.nan(table_mat)] <- 0
  print(table_mat)
}

# Data preprocessing ----

# Convert categorical variables to factors with labels
convert_factors <- function(df, list_of_vars_and_labels) {
  for (varname in names(list_of_vars_and_labels)) {
    df[[varname]] <- factor(df[[varname]], labels = list_of_vars_and_labels[[varname]])
  }
  return(df)
}

MENTAL <- convert_factors(MENTAL, list(
  PATHOLOGY = c("BD", "SCH"),
  SEX = c("Male", "Female"),
  HTT_CODE = c("NORMAL", "IA", "EXPANDED"),
  SCA1_CODE = c("NORMAL", "IA"),
  SCA2_CODE = c("NORMAL", "IA", "EXPANDED"),
  COFFEE = c("Non-coffee", "Coffee"),
  SMOKER = c("Smoking", "Non-smoking")
))

CONTROLS <- convert_factors(CONTROLS, list(
  PATHOLOGY = "CONTROL",
  SEX = c("Male", "Female"),
  HTT_CODE = c("NORMAL", "IA", "EXPANDED"),
  SCA1_CODE = c("NORMAL", "IA"),
  SCA2_CODE = c("NORMAL", "IA"),
  SMOKER = c("Smoking", "Non-smoking")
))

# Convert numeric variables
to_numeric <- function(df, vars) {
  df[vars] <- lapply(df[vars], as.numeric)
  return(df)
}

MENTAL <- to_numeric(MENTAL, c("ALLELE1_HTT", "ALLELE2_HTT", "ALLELE1_SCA1", "ALLELE2_SCA1",
                               "ALLELE1_SCA2", "ALLELE2_SCA2", "AGE", "ONSET_AGE", "DURATION"))

CONTROLS <- to_numeric(CONTROLS, c("ALLELE1_HTT", "ALLELE2_HTT", "ALLELE1_SCA1", "ALLELE2_SCA1",
                                   "ALLELE1_SCA2", "ALLELE2_SCA2", "AGE"))

# Subsetting by pathology groups ----

# Bipolar Disorder subset and transformations
BD <- subset(MENTAL, PATHOLOGY == "BD")
BD <- BD |> mutate(
  CD = factor(CD, labels = c("NO", "MILD", "MODERATE", "SEVERE", "VERY-SEVERE")),
  PATHOLOGY_TYPE = factor(PATHOLOGY_TYPE, labels = c("BD-I", "BD-II", "Cyclothymia", "Substance-related", "Other-BD")),
  CD_BINARY = factor(ifelse(CD == "NO", "NO", "CD"), levels = c("NO", "CD")),
  PATHOLOGY_TYPE_BINARY = factor(ifelse(PATHOLOGY_TYPE == "BD-I", "BD-I", "Other"), levels = c("BD-I", "Other"))
)

# Schizophrenia subset and transformations
SCH <- subset(MENTAL, PATHOLOGY == "SCH")
SCH <- SCH |> mutate(
  PATHOLOGY_TYPE_BINARY = factor(ifelse(PATHOLOGY_TYPE == 1, "SCH", "Other"), levels = c("SCH", "Other")),
  CD = factor(CD, labels = c("NO", "MILD", "MODERATE", "SEVERE")),
  CD_BINARY = factor(ifelse(CD == "NO", "NO", "CD"), levels = c("NO", "CD"))
)

# Add binary pathology columns to controls
CONTROLS$PATHOLOGY_TYPE_BINARY <- "CONTROL"
CONTROLS$CD_BINARY <- "CONTROL"

# Combine datasets ----

DT <- bind_rows(
  CONTROLS |> select(-CD_BINARY, -PATHOLOGY_TYPE_BINARY),
  MENTAL |> select(-COFFEE, -CD, -PATHOLOGY_TYPE, -ONSET_AGE, -DURATION)
)

DT$PATHOLOGY <- as.factor(DT$PATHOLOGY)

BD_CONTROLS <- bind_rows(
  BD |> select(-COFFEE, -CD, -ONSET_AGE, -DURATION, -PATHOLOGY_TYPE),
  CONTROLS
)

SCH_CONTROLS <- bind_rows(
  SCH |> select(-COFFEE,-CD, -ONSET_AGE, -DURATION, -PATHOLOGY_TYPE),
  CONTROLS
)

# Output section ----
# Save processed datasets for downstream analyses
if (!dir.exists("results")) {
  dir.create("results")
}

saveRDS(BD, file = "results/BD.rds")
saveRDS(SCH, file = "results/SCH.rds")
saveRDS(DT, file = "results/DT.rds")

# Save as XLSX
write_xlsx(BD, path = "results/BD.xlsx")
write_xlsx(SCH, path = "results/SCH.xlsx")
write_xlsx(DT, path = "results/DT.xlsx")
```

## 02_Demographic_analysis.R

This script performs descriptive comparisons across psychiatric groups (SCZ, BD, Controls) and genotype frequency analyses for *HTT*, *ATXN1* and *ATXN2* loci across subgroups (e.g., binary subtype, CD, controls). Analyses include demographic, clinical, lifestyle variables and STR genotype distributions.

```{r}
# Age at onset ----
mean_sd(MENTAL, "PATHOLOGY", "ONSET_AGE", "Age at onset")
wilcox.test(ONSET_AGE ~ PATHOLOGY, data = MENTAL, exact = TRUE)
rank_biserial(ONSET_AGE ~ PATHOLOGY, data = MENTAL)

# Age at death / last visit ----
mean_sd(DT, "PATHOLOGY", "AGE", "Age at death / last visit")
kruskal.test(AGE ~ PATHOLOGY, data = DT)
dunn <- dunnTest(DT$AGE ~ DT$PATHOLOGY, method = "holm")
print(dunn, dunn.test.results = TRUE)
pairwise.wilcox.test(x = DT$AGE, g = DT$PATHOLOGY, p.adjust.method = "holm")

# Rank biserial effect sizes by pairwise comparison
levels <- unique(DT$PATHOLOGY)
comparisons <- combn(levels, 2, simplify = FALSE)
for (pair in comparisons) {
  cat("\nComparison:", pair[1], "vs", pair[2], "\n")
  sub_data <- subset(DT, PATHOLOGY %in% pair)
  sub_data$PATHOLOGY <- factor(sub_data$PATHOLOGY, levels = pair)
  print(rank_biserial(AGE ~ PATHOLOGY, data = sub_data))
}

# Disease duration ----
mean_sd(MENTAL, "PATHOLOGY", "DURATION", "Disease duration")
wilcox.test(DURATION ~ PATHOLOGY, data = MENTAL, exact = TRUE)
pairwise.wilcox.test(x = MENTAL$DURATION, g = MENTAL$PATHOLOGY, p.adjust.method = "holm")
rank_biserial(DURATION ~ PATHOLOGY, data = MENTAL)


# Sex distribution ----
Table <- TABLE(DT, "PATHOLOGY", "SEX", "Sex distribution")
rowPercents(Table)
my_function(Table)
pairwise_fisher_test(Table, p.adjust.method = "holm", conf.int = TRUE, detailed = TRUE)

# Smoking status ----
smoking_tab <- TABLE(DT, "PATHOLOGY", "SMOKER", "Smoking by pathology")
rowPercents(smoking_tab)
my_function(smoking_tab)
pairwise_fisher_test(smoking_tab, p.adjust.method = "holm", conf.int = TRUE, detailed = TRUE)

# Coffee consumption ----
coffee_tab <- TABLE(MENTAL, "PATHOLOGY", "COFFEE", "Coffee consumption by pathology")
rowPercents(coffee_tab)
my_function(coffee_tab)
pairwise_fisher_test(coffee_tab, p.adjust.method = "holm", conf.int = TRUE, detailed = TRUE)



# Cognitive decline ----

cd_tab <- table(BD$CD)
cd_percent <- round(prop.table(cd_tab) * 100, 2)
cd_tab
cd_percent

cd_bin_tab <- table(BD$CD_BINARY)
round(prop.table(cd_bin_tab) * 100, 2)
cd_bin_tab

cd_tab <- table(SCH$CD)
cd_percent <- round(prop.table(cd_tab) * 100, 2)
cd_tab
cd_percent

cd_bin_tab <- table(SCH$CD_BINARY)
round(prop.table(cd_bin_tab) * 100, 2)
cd_bin_tab

# Subtype of pathology ----

subtype_tab <- table(BD$PATHOLOGY_TYPE)
round(prop.table(subtype_tab) * 100, 2)
subtype_tab
subtype_bin_tab <- table(BD$PATHOLOGY_TYPE_BINARY)
round(prop.table(subtype_bin_tab) * 100, 2)
subtype_bin_tab

sch_type_tab <- table(SCH$PATHOLOGY_TYPE)
round(prop.table(sch_type_tab) * 100, 2)
sch_type_tab
sch_type_bin_tab <- table(SCH$PATHOLOGY_TYPE_BINARY)
round(prop.table(sch_type_bin_tab) * 100, 2)
sch_type_bin_tab


# Missing data ----
DT %>%
  group_by(PATHOLOGY) %>%
  summarise(across(everything(), ~ sum(is.na(.))))

MENTAL %>%
  group_by(PATHOLOGY) %>%
  summarise(across(everything(), ~ sum(is.na(.))))


# Session info ----
sessionInfo()
```

## 03_Genotype_stats.R

This script performs genotype frequency comparisons across groups (SCZ, BD, Controls) for genes associated with neurodegenerative disorders: HTT, ATXN1 (SCA1), and ATXN2 (SCA2). The analysis includes global frequencies, intermediate alleles (IAs), expanded alleles, and subgroup comparisons by clinical subtype and severity (DCO).

```{r}
# Function to run genotype frequency analysis ----
run_genotype_analysis <- function(data, group_col, genotype_col, label, drop_cols = NULL) {
  tab <- TABLE(data, group_col, genotype_col, label)
  print(rowPercents(tab))
  res_misc <- my_function(tab)
  fisher_res <- NULL
  
  if (is.null(drop_cols)) {
    message("Pairwise Fisher skipped: requires 'drop_cols' with 2 categories.")
  } else {
    cols <- colnames(tab)
    idx_drop <- if (is.numeric(drop_cols)) intersect(drop_cols, seq_along(cols)) else match(drop_cols, cols, nomatch = 0)
    kept <- setdiff(seq_along(cols), idx_drop)
    mat <- as.matrix(tab[, kept, drop = FALSE])
    if (ncol(mat) == 2) {
      pairwise_fisher_test(mat, p.adjust.method = "holm", conf.int = TRUE, detailed = TRUE)
    } else {
      message("Pairwise Fisher skipped: reduced table not equal to 2 categories.")
    }
  }
}



# HTT ANALYSIS =================
run_genotype_analysis(DT, "PATHOLOGY", "HTT_CODE", "HTT: Main group comparison")
run_genotype_analysis(DT, "PATHOLOGY", "HTT_CODE", "HTT: Intermediate alleles", drop_cols = 3)
run_genotype_analysis(DT, "PATHOLOGY", "HTT_CODE", "HTT: Expanded alleles", drop_cols = 2)

run_genotype_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY	", "HTT_CODE", "HTT vs type of BD", drop_cols = 3)
run_genotype_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY	", "HTT_CODE", "HTT vs type of BD (expanded)", drop_cols = 2)

run_genotype_analysis(BD_CONTROLS, "CD_BINARY", "HTT_CODE", "HTT vs DCO severity in BD", drop_cols = 3)
run_genotype_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY	", "HTT_CODE", "HTT vs type of SCH", drop_cols = 3)
run_genotype_analysis(SCH_CONTROLS, "CD_BINARY", "HTT_CODE", "HTT vs DCO severity in SCH", drop_cols = 3)

# ATXN1 (SCA1) ANALYSIS =================
run_genotype_analysis(DT, "PATHOLOGY", "SCA1_CODE", "ATXN1: IA comparison")
run_genotype_analysis(DT, "PATHOLOGY", "SCA1_CODE", "ATXN1: Intermediate alleles", drop_cols = 3)
run_genotype_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY	", "SCA1_CODE", "SCA1 vs type of BP", drop_cols = 3)
run_genotype_analysis(BD_CONTROLS, "CD_BINARY", "SCA1_CODE", "SCA1 vs DCO severity in BP", drop_cols = 3)
run_genotype_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY	", "SCA1_CODE", "SCA1 vs type of SCH", drop_cols = 3)
run_genotype_analysis(SCH_CONTROLS, "CD_BINARY", "SCA1_CODE", "SCA1 vs DCO severity in SCH", drop_cols = 3)

# ATXN2 (SCA2) ANALYSIS =================
run_genotype_analysis(DT, "PATHOLOGY", "SCA2_CODE", "ATXN2: Main group comparison")
run_genotype_analysis(DT, "PATHOLOGY", "SCA2_CODE", "ATXN2: Intermediate alleles", drop_cols = 3)
run_genotype_analysis(DT, "PATHOLOGY", "SCA2_CODE", "ATXN2: Expanded alleles", drop_cols = 2)

run_genotype_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY	", "SCA2_CODE", "SCA2 vs type of BP", drop_cols = 3)
run_genotype_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY	", "SCA2_CODE", "SCA2 vs type of BD (expanded)", drop_cols = 2)

run_genotype_analysis(BD_CONTROLS, "CD_BINARY", "SCA2_CODE", "SCA2 vs DCO severity in BP", drop_cols = 3)
run_genotype_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY	", "SCA2_CODE", "SCA2 vs type of SCH", drop_cols = 3)
run_genotype_analysis(SCH_CONTROLS, "CD_BINARY", "SCA2_CODE", "SCA2 vs DCO severity in SCH", drop_cols = 3)


# Session info ----
sessionInfo()
```

## 04_CAG_repeat_sizes.R

This script performs non-parametric statistical analyses to compare CAG repeat lengths of both alleles in *HTT*, *ATXN1*, and *ATXN2* across diagnostic groups (SCZ, BD, Controls), as well as in clinically defined subgroups based on subtype and severity (DCO).

The analysis includes:

-   Descriptive statistics (mean and SD per group)

-   Kruskal-Wallis test for global group differences

-   Dunn and Wilcoxon pairwise post-hoc tests if significant

-   Effect size estimation (Wilcoxon-based) for selected comparisons

-   CAG repeat sizes and its percentages in the different cohorts.

```{r}
run_kruskal_analysis <- function(df, group_col, value_col, result_name = NULL, comparisons_list = NULL) {
  # If result_name is not provided, use the name of value_col
  if (is.null(result_name)) {
    result_name <- value_col
  }
  
  cat("\n--- Analysis for:", result_name, "---\n")
  
  # 1. Mean and standard deviation
  cat("Mean and SD by group:\n")
  mean_sd_result <- mean_sd(df, group_col, value_col, result_name = value_col)
  print(mean_sd_result)
  
  # 2. Kruskal-Wallis test
  cat("\nKruskal-Wallis test:\n")
  kruskal <- kruskal.test(df[[value_col]] ~ df[[group_col]])
  print(kruskal)
  
  # 3. If significant, perform post-hoc tests and effect size
  if (kruskal$p.value < 0.05) {
    cat("\nDunn's test (Holm correction):\n")
    dunn_result <- FSA::dunnTest(df[[value_col]] ~ df[[group_col]], method = "holm")
    print(dunn_result)
    
    cat("\nPairwise Wilcoxon test (Holm correction):\n")
    wilcox_result <- pairwise.wilcox.test(df[[value_col]], df[[group_col]], p.adjust.method = "holm")
    print(wilcox_result)
    
    # 4. Effect size using wilcox_effsize if comparisons provided
    if (!is.null(comparisons_list)) {
      cat("\nEffect size for selected comparisons:\n")
      effsize <- df %>%
        wilcox_effsize(as.formula(paste(value_col, "~", group_col)),
                       paired = FALSE,
                       comparisons = comparisons_list,
                       p.adjust.method = "holm")
      print(effsize)
    } else {
      cat("No specific comparisons provided for effect size.\n")
    }
  } else {
    cat("\nKruskal-Wallis not significant; post-hoc tests and effect sizes are not performed.\n")
  }
}


# HTT ANALYSIS ----

### Long allele
run_kruskal_analysis(DT, "PATHOLOGY", "ALLELE2_HTT", "HTT long allele", list(c("BD","SCH"), c("BD","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE2_HTT", "HTT long allele BD_CONTROLS - Pathology type", list(c("TBD-1","Other"), c("TBD-1","CONTROL"), c("Other","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "CD_BINARY", "ALLELE2_HTT", "HTT long allele BD_CONTROLS - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE2_HTT", "HTT long allele SCH_CONTROLS - Pathology type", list(c("SCH","Other"), c("Other","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "CD_BINARY", "ALLELE2_HTT", "HTT long allele SCH_CONTROLS - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))

### Short allele
run_kruskal_analysis(DT, "PATHOLOGY", "ALLELE1_HTT", "HTT short allele", list(c("BD","SCH"), c("BD","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE1_HTT", "HTT short allele BD_CONTROLES - Pathology type", list(c("TBD-1","Other"), c("TBD-1","CONTROL"), c("Other","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "CD_BINARY", "ALLELE1_HTT", "HTT short allele BD_CONTROLES - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE1_HTT", "HTT short allele SCH_CONTROLS - Pathology type", list(c("SCH","Other"), c("Other","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "CD_BINARY", "ALLELE1_HTT", "HTT short allele SCH_CONTROLS - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))

# ATXN1 (SCA1) ANALYSIS ----

### Long allele
run_kruskal_analysis(DT, "PATHOLOGY", "ALLELE2_SCA1", "SCA1 long allele", list(c("BD","SCH"), c("BD","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE2_SCA1", "SCA1 long allele BD_CONTROLES - Pathology type", list(c("TBD-1","Other"), c("TBD-1","CONTROL"), c("Other","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "CD_BINARY", "ALLELE2_SCA1", "SCA1 long allele BD_CONTROLES - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE2_SCA1", "SCA1 long allele SCH_CONTROLS - Pathology type", list(c("SCH","Other"), c("Other","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "CD_BINARY", "ALLELE2_SCA1", "SCA1 long allele SCH_CONTROLS - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))

### Short allele
run_kruskal_analysis(DT, "PATHOLOGY", "ALLELE1_SCA1", "SCA1 short allele", list(c("BD","SCH"), c("BD","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE1_SCA1", "SCA1 short allele BD_CONTROLES - Pathology type", list(c("TBD-1","Other"), c("TBD-1","CONTROL"), c("Other","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "CD_BINARY", "ALLELE1_SCA1", "SCA1 short allele BD_CONTROLES - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE1_SCA1", "SCA1 short allele SCH_CONTROLS - Pathology type", list(c("SCH","Other"), c("Other","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "CD_BINARY", "ALLELE1_SCA1", "SCA1 short allele SCH_CONTROLS - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))

# ATXN2 (SCA2) ANALYSIS ----

### Long allele
run_kruskal_analysis(DT, "PATHOLOGY", "ALLELE2_SCA2", "SCA2 long allele", list(c("BD","SCH"), c("BD","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE2_SCA2", "SCA2 long allele BD_CONTROLES - Pathology type", list(c("BD-I","Other"), c("BD-I","CONTROL"), c("Other","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "CD_BINARY", "ALLELE2_SCA2", "SCA2 long allele BD_CONTROLES - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE2_SCA2", "SCA2 long allele SCH_CONTROLS - Pathology type", list(c("SCH","Other"), c("Other","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "CD_BINARY", "ALLELE2_SCA2", "SCA2 long allele SCH_CONTROLS - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))

### Short allele
run_kruskal_analysis(DT, "PATHOLOGY", "ALLELE1_SCA2", "SCA2 short allele", list(c("BD","SCH"), c("BD","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE1_SCA2", "SCA2 short allele BD_CONTROLES - Pathology type", list(c("TBD-1","Other"), c("TBD-1","CONTROL"), c("Other","CONTROL")))
run_kruskal_analysis(BD_CONTROLS, "CD_BINARY", "ALLELE1_SCA2", "SCA2 short allele BD_CONTROLES - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "PATHOLOGY_TYPE_BINARY", "ALLELE1_SCA2", "SCA2 short allele SCH_CONTROLS - Pathology type", list(c("SCH","Other"), c("Other","CONTROL"), c("SCH","CONTROL")))
run_kruskal_analysis(SCH_CONTROLS, "CD_BINARY", "ALLELE1_SCA2", "SCA2 short allele SCH_CONTROLS - Severity", list(c("CD","NO"), c("CD","CONTROL"), c("NO","CONTROL")))

#CAG repeats frequencies among cohorts----
library(dplyr)
library(tidyr)
library(rlang)

allele_percentages_by_group <- function(
    data,
    group_var      = "PATHOLOGY",
    allele_cols    = c("ALLELE1_HTT", "ALLELE2_HTT"),  # <- pass SCA1/SCA2 here when needed
    group_to_show  = NULL,                              # e.g. "BD", "SCH", "CONTROL"
    output         = c("long", "wide"),                 # output format
    digits         = 2,
    drop_na        = TRUE
) {
  output <- match.arg(output)
  
  # Check that all requested columns exist in the dataset
  missing_cols <- setdiff(c(group_var, allele_cols), names(data))
  if (length(missing_cols) > 0) {
    stop("These columns were not found in 'data': ", paste(missing_cols, collapse = ", "))
  }
  
  # Convert selected allele columns to long format
  allele_long <- data %>%
    select(all_of(group_var), all_of(allele_cols)) %>%
    pivot_longer(
      cols      = all_of(allele_cols),
      names_to  = "ALLELE_COL",
      values_to = "CAG_size"
    )
  
  # Optionally remove rows with missing allele values
  if (drop_na) {
    allele_long <- allele_long %>% filter(!is.na(CAG_size))
  }
  
  # Compute frequency and percentage by group and allele size
  allele_freq <- allele_long %>%
    group_by(across(all_of(group_var)), CAG_size) %>%
    summarise(count = n(), .groups = "drop_last") %>%
    mutate(
      percentage = round(100 * count / sum(count), digits)  # % within each group
    ) %>%
    arrange(across(all_of(group_var)), CAG_size)
  
  # Optionally filter to show only one specific group
  if (!is.null(group_to_show)) {
    allele_freq <- allele_freq %>%
      filter(!!sym(group_var) == group_to_show)
  }
  
  # Return results in wide format if requested
  if (output == "wide") {
    res <- allele_freq %>%
      select(all_of(group_var), CAG_size, percentage) %>%
      pivot_wider(
        names_from  = all_of(group_var),
        values_from = percentage,
        values_fill = 0
      ) %>%
      arrange(CAG_size)
  } else {
    res <- allele_freq
  }
  
  # Print and return the result invisibly
  print(res)
  invisible(res)
}

allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_HTT", "ALLELE2_HTT"),  group_to_show = "CONTROL")
allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_HTT", "ALLELE2_HTT"),  group_to_show = "BD")
allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_HTT", "ALLELE2_HTT"),  group_to_show = "SCH")

allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_SCA1", "ALLELE2_SCA1"),  group_to_show = "CONTROL")
allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_SCA1", "ALLELE2_SCA1"),  group_to_show = "BD")
allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_SCA1", "ALLELE2_SCA1"),  group_to_show = "SCH")

allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_SCA2", "ALLELE2_SCA2"),  group_to_show = "CONTROL")
allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_SCA2", "ALLELE2_SCA2"),  group_to_show = "BD")
allele_percentages_by_group(DT,  allele_cols = c("ALLELE1_SCA2", "ALLELE2_SCA2"),  group_to_show = "SCH")

```

## 05_Regression_models.R

This script performs regression analyses to investigate:

1)  Association between STR genotypes (*HTT*, *ATXN1*, *ATXN2*) and:

\- Risk of psychiatric disorders (BD, SCZ vs controls)

Risk of clinical subtypes (e.g., bipolar subtypes)

\- Risk of cognitive deterioration (CD)

2)  Association between STRs and age-of-onset in psychiatric disorders Includes covariate adjustment, model selection, and effect size calculations.

```{r}
# MENTAL DISORDER RISK MODEL----

# Multinomial model (3 categories)
DT$PATHOLOGY <- relevel(DT$PATHOLOGY, ref = "CONTROL")

modelo_multi <- multinom(PATHOLOGY ~ ., data = na.omit(DT))
summary(modelo_multi)
step(modelo_multi, direction = "backward", trace = 0)

modelo_multi <- multinom(PATHOLOGY ~ SEX + AGE + SMOKER + ALLELE2_SCA1 + 
                           SCA1_CODE + ALLELE2_SCA2 + SCA2_CODE, 
                         data = na.omit(DT))
summary(modelo_multi)

# Calculate p-values
z <- summary(modelo_multi)$coefficients / summary(modelo_multi)$standard.errors
pvals <- 2 * (1 - pnorm(abs(z)))
pvals
exp(coef(modelo_multi))
confint(modelo_multi)

# Binomial model (2 categories)
MENTAL$PATHOLOGY <- relevel(MENTAL$PATHOLOGY, ref = "BD")
MLM <- select(MENTAL, -ONSET_AGE, -AGE, -DURATION, -PATHOLOGY_TYPE, -CD)

# Full model 
model <- glm(PATHOLOGY ~ ., family = binomial, data = na.omit(MLM))
summary(model)

# Backward selection
step(model, direction = "backward", trace = 0)

# Final model
mod <- glm(PATHOLOGY ~ SEX + SCA1_CODE + ALLELE2_SCA2 + SCA2_CODE, 
           data = MLM, family = binomial)
summary(mod)
lrtest(mod)
confint(mod)
exp(coef(mod))

# HTT RISK MODEL ----
DTLM <- MENTAL %>%
  filter(!is.na(ALLELE1_HTT), !is.na(ALLELE2_HTT), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER))

# Full model with covariates
model <- glm(PATHOLOGY ~ SEX + COFFEE + SMOKER + 
               ALLELE1_HTT + I(ALLELE1_HTT^2) + 
               ALLELE2_HTT + I(ALLELE2_HTT^2) + 
               ALLELE1_HTT:ALLELE2_HTT,
             data = DTLM, family = binomial)
summary(model)
step(model, direction = "backward", trace = 0)

# Final model
mod <- glm(PATHOLOGY ~ SEX, 
           data = MENTAL, family = binomial)
summary(mod)
lrtest(mod)
confint(mod)
exp(coef(mod))

# Model without covariates
model_nocov <- glm(PATHOLOGY ~ ALLELE1_HTT + I(ALLELE1_HTT^2) + 
                     ALLELE2_HTT + I(ALLELE2_HTT^2) + 
                     ALLELE1_HTT:ALLELE2_HTT,
                   data = DTLM, family = binomial)
summary(model_nocov)
summary(step(model_nocov, direction = "backward", trace = 0))

# Cognitive Deterioration (CD) model----
BD$CD_BINARY <- relevel(BD$CD_BINARY, ref = "CD")
BDLM <- BD %>%
  filter(!is.na(ALLELE1_HTT), !is.na(ALLELE2_HTT), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(CD_BINARY))

model_CD <- glm(CD_BINARY ~ SEX + COFFEE + SMOKER + 
                  ALLELE1_HTT + I(ALLELE1_HTT^2) + 
                  ALLELE2_HTT + I(ALLELE2_HTT^2) + 
                  ALLELE1_HTT:ALLELE2_HTT,
                data = BDLM, family = binomial())
summary(model_CD)
summary(step(model_CD, direction = "backward", trace = 0))

SCH$CD_BINARY <- relevel(SCH$CD_BINARY, ref = "CD")
SCHLM <- SCH %>%
  filter(!is.na(ALLELE1_HTT), !is.na(ALLELE2_HTT), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(CD_BINARY))

model_CD <- glm(CD_BINARY ~ SEX + COFFEE + SMOKER + 
                  ALLELE1_HTT + I(ALLELE1_HTT^2) + 
                  ALLELE2_HTT + I(ALLELE2_HTT^2) + 
                  ALLELE1_HTT:ALLELE2_HTT,
                data = SCHLM, family = binomial())
summary(model_CD)
summary(step(model_CD, direction = "backward", trace = 0))

#Subtypes----
# Bipolar Subtype
BDLM <- BD %>%
  filter(!is.na(ALLELE1_HTT), !is.na(ALLELE2_HTT), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(PATHOLOGY_TYPE_BINARY))

model_BD <- glm(PATHOLOGY_TYPE_BINARY ~ SEX + COFFEE + SMOKER + 
                  ALLELE1_HTT + I(ALLELE1_HTT^2) + 
                  ALLELE2_HTT + I(ALLELE2_HTT^2) + 
                  ALLELE1_HTT:ALLELE2_HTT,
                data = BDLM, family = binomial())
summary(model_BD)
summary(step(model_BD, direction = "backward", trace = 0))

# Schizophrenia Subtype
SCHLM <- SCH %>%
  filter(!is.na(ALLELE1_HTT), !is.na(ALLELE2_HTT), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(PATHOLOGY_TYPE_BINARY))
model_SCH <- glm(PATHOLOGY_TYPE_BINARY ~ ALLELE1_HTT + I(ALLELE1_HTT^2) + 
                   ALLELE2_HTT + I(ALLELE2_HTT^2) + 
                   ALLELE1_HTT:ALLELE2_HTT,
                 data = SCHLM, family = binomial())
summary(model_SCH)
summary(step(model_SCH, direction = "backward", trace = 0))


# ATXN1 RISK MODEL ----
DTLM <- MENTAL %>%
  filter(!is.na(ALLELE1_SCA1), !is.na(ALLELE2_SCA1), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER))

# Full model with covariates
model_ATXN1 <- glm(PATHOLOGY ~ SEX + COFFEE + SMOKER + 
                     ALLELE1_SCA1 + I(ALLELE1_SCA1^2) + 
                     ALLELE2_SCA1 + I(ALLELE2_SCA1^2) + 
                     ALLELE1_SCA1:ALLELE2_SCA1,
                   data = DTLM, family = binomial)
summary(model_ATXN1)
summary(step(model_ATXN1, direction = "backward", trace = 0))

# Final model
mod_ATXN1 <- glm(PATHOLOGY ~ SEX + ALLELE1_SCA1 + ALLELE2_SCA1 + 
                   ALLELE1_SCA1:ALLELE2_SCA1, 
                 data = MENTAL, family = binomial)
summary(mod_ATXN1)
lrtest(mod_ATXN1)
confint(mod_ATXN1)
exp(coef(mod_ATXN1))

# Model without covariates
model_ATXN1_nocov <- glm(PATHOLOGY ~ ALLELE1_SCA1 + I(ALLELE1_SCA1^2) + 
                           ALLELE2_SCA1 + I(ALLELE2_SCA1^2) + 
                           ALLELE1_SCA1:ALLELE2_SCA1,
                         data = DTLM, family = binomial)
summary(model_ATXN1_nocov)
summary(step(model_ATXN1_nocov, direction = "backward", trace = 0))

# Cognitive Deterioration (CD) model----
BDLM_ATXN1 <- BD %>%
  filter(!is.na(ALLELE1_SCA1), !is.na(ALLELE2_SCA1), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(CD_BINARY))

model_CD_ATXN1 <- glm(CD_BINARY ~ SEX + COFFEE + SMOKER + 
                        ALLELE1_SCA1 + I(ALLELE1_SCA1^2) + 
                        ALLELE2_SCA1 + I(ALLELE2_SCA1^2) + 
                        ALLELE1_SCA1:ALLELE2_SCA1,
                      data = BDLM_ATXN1, family = binomial())
summary(model_CD_ATXN1)
summary(step(model_CD_ATXN1, direction = "backward", trace = 0))

SCHLM_ATXN1 <- SCH %>%
  filter(!is.na(ALLELE1_SCA1), !is.na(ALLELE2_SCA1), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(CD_BINARY))

model_CD_ATXN1 <- glm(CD_BINARY ~ SEX + COFFEE + SMOKER + 
                        ALLELE1_SCA1 + I(ALLELE1_SCA1^2) + 
                        ALLELE2_SCA1 + I(ALLELE2_SCA1^2) + 
                        ALLELE1_SCA1:ALLELE2_SCA1,
                      data = SCHLM_ATXN1, family = binomial())
summary(model_CD_ATXN1)
summary(step(model_CD_ATXN1, direction = "backward", trace = 0))
#Subtypes----
# Bipolar Subtype
BDLM_ATXN1 <- BD %>%
  filter(!is.na(ALLELE1_SCA1), !is.na(ALLELE2_SCA1), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(PATHOLOGY_TYPE_BINARY))

model_BD_ATXN1 <- glm(PATHOLOGY_TYPE_BINARY ~ SEX + COFFEE + SMOKER + 
                        ALLELE1_SCA1 + I(ALLELE1_SCA1^2) + 
                        ALLELE2_SCA1 + I(ALLELE2_SCA1^2) + 
                        ALLELE1_SCA1:ALLELE2_SCA1,
                      data = BDLM_ATXN1, family = binomial())
summary(model_BD_ATXN1)
summary(step(model_BD_ATXN1, direction = "backward", trace = 0))

# Schizophrenia Subtype
SCHLM_ATXN1 <- SCH %>%
  filter(!is.na(ALLELE1_SCA1), !is.na(ALLELE2_SCA1), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(PATHOLOGY_TYPE_BINARY))

model_SCH_ATXN1 <- glm(PATHOLOGY_TYPE_BINARY ~ SEX + COFFEE + SMOKER + 
                         ALLELE1_SCA1 + I(ALLELE1_SCA1^2) + 
                         ALLELE2_SCA1 + I(ALLELE2_SCA1^2) + 
                         ALLELE1_SCA1:ALLELE2_SCA1,
                       data = SCHLM_ATXN1, family = binomial())
summary(model_SCH_ATXN1)
summary(step(model_SCH_ATXN1, direction = "backward", trace = 0))


# ATXN2 RISK MODEL ----
DTLM_ATXN2 <- MENTAL %>%
  filter(!is.na(ALLELE1_SCA2), !is.na(ALLELE2_SCA2), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER))

# Full model with covariates
model_ATXN2 <- glm(PATHOLOGY ~ SEX + COFFEE + SMOKER + 
                     ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                     ALLELE2_SCA2 + I(ALLELE2_SCA2^2) + 
                     ALLELE1_SCA2:ALLELE2_SCA2,
                   data = DTLM_ATXN2, family = binomial)
summary(model_ATXN2)
summary(step(model_ATXN2, direction = "backward", trace = 0))

# Final model
mod_ATXN2 <- glm(PATHOLOGY ~ SEX + ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                   ALLELE2_SCA2 + I(ALLELE2_SCA2^2),
                 data = MENTAL, family = binomial)
summary(mod_ATXN2)
lrtest(mod_ATXN2)
confint(mod_ATXN2)
exp(coef(mod_ATXN2))
summary(step(mod_ATXN2, direction = "backward", trace = 0))

# Model without covariates
model_ATXN2_nocov <- glm(PATHOLOGY ~ ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                           ALLELE2_SCA2 + I(ALLELE2_SCA2^2) + 
                           ALLELE1_SCA2:ALLELE2_SCA2,
                         data = DTLM_ATXN2, family = binomial)
summary(model_ATXN2_nocov)
summary(step(model_ATXN2_nocov, direction = "backward", trace = 0))

# Cognitive Deterioration (CD) model----
BDLM_ATXN2 <- BD %>%
  filter(!is.na(ALLELE1_SCA2), !is.na(ALLELE2_SCA2), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(CD_BINARY))

model_CD_ATXN2 <- glm(CD_BINARY ~ SEX + COFFEE + SMOKER + 
                        ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                        ALLELE2_SCA2 + I(ALLELE2_SCA2^2) + 
                        ALLELE1_SCA2:ALLELE2_SCA2,
                      data = BDLM_ATXN2, family = binomial())
summary(model_CD_ATXN2)
summary(step(model_CD_ATXN2, direction = "backward", trace = 0))

SCHLM_ATXN2 <- SCH %>%
  filter(!is.na(ALLELE1_SCA2), !is.na(ALLELE2_SCA2), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(CD_BINARY))

model_CD_ATXN2 <- glm(CD_BINARY ~ SEX + COFFEE + SMOKER + 
                        ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                        ALLELE2_SCA2 + I(ALLELE2_SCA2^2) + 
                        ALLELE1_SCA2:ALLELE2_SCA2,
                      data = SCHLM_ATXN2, family = binomial())
summary(model_CD_ATXN2)
summary(step(model_CD_ATXN2, direction = "backward", trace = 0))
#Subtypes----
# Bipolar Subtype
BDLM_ATXN2 <- BD %>%
  filter(!is.na(ALLELE1_SCA2), !is.na(ALLELE2_SCA2), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(PATHOLOGY_TYPE_BINARY))

model_BD_ATXN2 <- glm(PATHOLOGY_TYPE_BINARY ~ SEX + COFFEE + SMOKER + 
                        ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                        ALLELE2_SCA2 + I(ALLELE2_SCA2^2) + 
                        ALLELE1_SCA2:ALLELE2_SCA2,
                      data = BDLM_ATXN2, family = binomial())
summary(model_BD_ATXN2)
summary(step(model_BD_ATXN2, direction = "backward", trace = 0))

# Schizophrenia Subtype
SCHLM_ATXN2 <- SCH %>%
  filter(!is.na(ALLELE1_SCA2), !is.na(ALLELE2_SCA2), !is.na(SEX), 
         !is.na(COFFEE), !is.na(SMOKER), !is.na(PATHOLOGY_TYPE_BINARY))

model_SCH_ATXN2 <- glm(PATHOLOGY_TYPE_BINARY ~ SEX + COFFEE + SMOKER + 
                         ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                         ALLELE2_SCA2 + I(ALLELE2_SCA2^2) + 
                         ALLELE1_SCA2:ALLELE2_SCA2,
                       data = SCHLM_ATXN2, family = binomial())
summary(model_SCH_ATXN2)
summary(step(model_SCH_ATXN2, direction = "backward", trace = 0))



# ==============================================================================

# AGE-ONSET MODELING ----


# HTT and Age-of-Onset ----

# Bipolar disorder
model_onset_BD <- lm(ONSET_AGE ~ SEX + COFFEE + SMOKER + 
                       ALLELE1_HTT + I(ALLELE1_HTT^2) + 
                       ALLELE2_HTT + I(ALLELE2_HTT^2) + 
                       ALLELE1_HTT:ALLELE2_HTT, 
                     data = na.omit(BD))
summary(model_onset_BD)
summary(step(model_onset_BD, direction = "backward", trace = 0))

# Schizophrenia
model_onset_SCH <- lm(ONSET_AGE ~ SEX + COFFEE + SMOKER + 
                        ALLELE1_HTT + I(ALLELE1_HTT^2) + 
                        ALLELE2_HTT + I(ALLELE2_HTT^2) + 
                        ALLELE1_HTT:ALLELE2_HTT, 
                      data = na.omit(SCH))
summary(model_onset_SCH)
summary(step(model_onset_SCH, direction = "backward", trace = 0))

# ATXN1 and Age-of-Onset ----

# Bipolar Disorder
model_onset_BD_ATXN1 <- lm(ONSET_AGE ~ SEX + COFFEE + SMOKER + 
                             ALLELE1_SCA1 + I(ALLELE1_SCA1^2) + 
                             ALLELE2_SCA1 + I(ALLELE2_SCA1^2) + 
                             ALLELE1_SCA1:ALLELE2_SCA1,
                           data = na.omit(BD))
summary(model_onset_BD_ATXN1)
summary(step(model_onset_BD_ATXN1, direction = "backward", trace = 0))

# Schizophrenia
model_onset_SCH_ATXN1 <- lm(ONSET_AGE ~ SEX + COFFEE + SMOKER + 
                              ALLELE1_SCA1 + I(ALLELE1_SCA1^2) + 
                              ALLELE2_SCA1 + I(ALLELE2_SCA1^2) + 
                              ALLELE1_SCA1:ALLELE2_SCA1,
                            data = na.omit(SCH))
summary(model_onset_SCH_ATXN1)
summary(step(model_onset_SCH_ATXN1, direction = "backward", trace = 0))


# ATXN2 and Age-of-Onset ----

# Bipolar Disorder
model_onset_BD_ATXN2 <- lm(ONSET_AGE ~ SEX + COFFEE + SMOKER + 
                             ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                             ALLELE2_SCA2 + I(ALLELE2_SCA2^2) + 
                             ALLELE1_SCA2:ALLELE2_SCA2,
                           data = na.omit(BD))
summary(model_onset_BD_ATXN2)
summary(step(model_onset_BD_ATXN2, direction = "backward", trace = 0))

# Schizophrenia
model_onset_SCH_ATXN2 <- lm(ONSET_AGE ~ SEX + COFFEE + SMOKER + 
                              ALLELE1_SCA2 + I(ALLELE1_SCA2^2) + 
                              ALLELE2_SCA2 + I(ALLELE2_SCA2^2) + 
                              ALLELE1_SCA2:ALLELE2_SCA2,
                            data = na.omit(SCH))
summary(model_onset_SCH_ATXN2)
summary(step(model_onset_SCH_ATXN2, direction = "backward", trace = 0))
```

## 06_Survival_Age_analysis.R

This script evaluates the association between STR-based genotype classifications (HTT_CODE, SCA1_CODE, SCA2_CODE) and clinical variables (age at onset, disease duration) in schizophrenia (SCH) and bipolar disorder (BD) patients. Analyses include subgroup comparisons within BD (e.g., BD-I, BD-CD) and survival models.

```{r}
# Subsetting by BD-I and CD
BD_I <- subset(BD, BD$PATHOLOGY_TYPE_BINARY == "BD-I")
BD_CD <- subset(BD, BD$CD_BINARY == "CD")
BD_NOCD <- subset(BD, BD$CD_BINARY == "NO")

# Subsetting by CD in SCH
SCH_CD <- subset(SCH, SCH$CD_BINARY == "CD")
SCH_NOCD <- subset(SCH, SCH$CD_BINARY == "NO")

# AGE AT ONSET CORRELATIONS ----

# BD - HTT
mean_sd(BD, "HTT_CODE", "ONSET_AGE", "Age at onset BD")
kruskal.test(ONSET_AGE ~ HTT_CODE, data = BD)
dunn <- dunnTest(BD$ONSET_AGE ~ BD$HTT_CODE, method = "holm")
print(dunn, dunn.test.results = TRUE)

mean_sd(BD_I, "HTT_CODE", "ONSET_AGE", "Age at onset BD-I")
kruskal.test(ONSET_AGE ~ HTT_CODE, data = BD_I)

mean_sd(BD_CD, "HTT_CODE", "ONSET_AGE", "Age at onset BD-CD")
kruskal.test(ONSET_AGE ~ HTT_CODE, data = BD_CD)

mean_sd(BD_NOCD, "HTT_CODE", "ONSET_AGE", "Age at onset BD-NO-CD")
kruskal.test(ONSET_AGE ~ HTT_CODE, data = BD_NOCD)

# SCH - HTT
mean_sd(SCH, "HTT_CODE", "ONSET_AGE", "Age at onset SCH")
wilcox.test(ONSET_AGE ~ HTT_CODE, data = SCH)

mean_sd(SCH_CD, "HTT_CODE", "ONSET_AGE", "Age at onset SCH_CD")
kruskal.test(ONSET_AGE ~ HTT_CODE, data = SCH_CD)

mean_sd(SCH_NOCD, "HTT_CODE", "ONSET_AGE", "Age at onset SCH_NOCD")
kruskal.test(ONSET_AGE ~ HTT_CODE, data = SCH_NOCD)

# BD - SCA1
mean_sd(BD, "SCA1_CODE", "ONSET_AGE", "Age at onset BD")
wilcox.test(ONSET_AGE ~ SCA1_CODE, data = BD)

mean_sd(BD_I, "SCA1_CODE", "ONSET_AGE", "Age at onset BD-I")
kruskal.test(ONSET_AGE ~ SCA1_CODE, data = BD_I)

mean_sd(BD_CD, "SCA1_CODE", "ONSET_AGE", "Age at onset BD-CD")
kruskal.test(ONSET_AGE ~ SCA1_CODE, data = BD_CD)

mean_sd(BD_NOCD, "SCA1_CODE", "ONSET_AGE", "Age at onset BD-NO-CD")
kruskal.test(ONSET_AGE ~ SCA1_CODE, data = BD_NOCD)

# SCH - SCA1
mean_sd(SCH, "SCA1_CODE", "ONSET_AGE", "Age at onset SCH")
wilcox.test(ONSET_AGE ~ SCA1_CODE, data = SCH)

mean_sd(SCH_CD, "SCA1_CODE", "ONSET_AGE", "Age at onset SCH_CD")
kruskal.test(ONSET_AGE ~ SCA1_CODE, data = SCH_CD)

mean_sd(SCH_NOCD, "SCA1_CODE", "ONSET_AGE", "Age at onset SCH_NOCD")
kruskal.test(ONSET_AGE ~ SCA1_CODE, data = SCH_NOCD)

# BD - SCA2
mean_sd(BD, "SCA2_CODE", "ONSET_AGE", "Age at onset BD")
wilcox.test(ONSET_AGE ~ SCA2_CODE, data = BD)

mean_sd(BD_I, "SCA2_CODE", "ONSET_AGE", "Age at onset BD-I")
wilcox.test(ONSET_AGE ~ SCA2_CODE, data = BD_I)

mean_sd(BD_CD, "SCA2_CODE", "ONSET_AGE", "Age at onset BD-CD")
wilcox.test(ONSET_AGE ~ SCA2_CODE, data = BD_CD)

mean_sd(BD_NOCD, "SCA2_CODE", "ONSET_AGE", "Age at onset BD-NO-CD")
kruskal.test(ONSET_AGE ~ SCA2_CODE, data = BD_NOCD)

# SCH - SCA2
mean_sd(SCH, "SCA2_CODE", "ONSET_AGE", "Age at onset SCH")
kruskal.test(ONSET_AGE ~ SCA2_CODE, data = SCH)

mean_sd(SCH_CD, "SCA2_CODE", "ONSET_AGE", "Age at onset SCH_CD")
kruskal.test(ONSET_AGE ~ SCA2_CODE, data = SCH_CD)

mean_sd(SCH_NOCD, "SCA2_CODE", "ONSET_AGE", "Age at onset SCH_NOCD")
kruskal.test(ONSET_AGE ~ SCA2_CODE, data = SCH_NOCD)

# DURATION CORRELATIONS ----

# BD - HTT
mean_sd(BD, "HTT_CODE", "DURATION", "Duration BD")
kruskal.test(DURATION ~ HTT_CODE, data = BD)

mean_sd(BD_I, "HTT_CODE", "DURATION", "Duration BD-I")
kruskal.test(DURATION ~ HTT_CODE, data = BD_I)

mean_sd(BD_CD, "HTT_CODE", "DURATION", "Duration BD-CD")
kruskal.test(DURATION ~ HTT_CODE, data = BD_CD)

mean_sd(BD_NOCD, "HTT_CODE", "DURATION", "Duration BD-NO-CD")
kruskal.test(DURATION ~ HTT_CODE, data = BD_NOCD)

# SCH - HTT
mean_sd(SCH, "HTT_CODE", "DURATION", "Duration SCH")
wilcox.test(DURATION ~ HTT_CODE, data = SCH)

mean_sd(SCH_CD, "HTT_CODE", "DURATION", "Duration SCH-CD")
kruskal.test(DURATION ~ HTT_CODE, data = SCH_CD)

mean_sd(SCH_NOCD, "HTT_CODE", "DURATION", "Duration SCH-NO-CD")
kruskal.test(DURATION ~ HTT_CODE, data = SCH_NOCD)

# BD - SCA1
mean_sd(BD, "SCA1_CODE", "DURATION", "Duration BD")
wilcox.test(DURATION ~ SCA1_CODE, data = BD)

mean_sd(BD_I, "SCA1_CODE", "DURATION", "Duration BD-I")
kruskal.test(DURATION ~ SCA1_CODE, data = BD_I)

mean_sd(BD_CD, "SCA1_CODE", "DURATION", "Duration BD-CD")
kruskal.test(DURATION ~ SCA1_CODE, data = BD_CD)

mean_sd(BD_NOCD, "SCA1_CODE", "DURATION", "Duration BD-NO-CD")
kruskal.test(DURATION ~ SCA1_CODE, data = BD_NOCD)

# SCH - SCA1
mean_sd(SCH, "SCA1_CODE", "DURATION", "Duration SCH")
wilcox.test(DURATION ~ SCA1_CODE, data = SCH)

mean_sd(SCH_CD, "SCA1_CODE", "DURATION", "Duration SCH-CD")
kruskal.test(DURATION ~ SCA1_CODE, data = SCH_CD)

mean_sd(SCH_NOCD, "SCA1_CODE", "DURATION", "Duration SCH-NO-CD")
kruskal.test(DURATION ~ SCA1_CODE, data = SCH_NOCD)

# BD - SCA2
mean_sd(BD, "SCA2_CODE", "DURATION", "Duration BD")
wilcox.test(DURATION ~ SCA2_CODE, data = BD)

mean_sd(BD_I, "SCA2_CODE", "DURATION", "Duration BD-I")
wilcox.test(DURATION ~ SCA2_CODE, data = BD_I)

mean_sd(BD_CD, "SCA2_CODE", "DURATION", "Duration BD-CD")
wilcox.test(DURATION ~ SCA2_CODE, data = BD_CD)

mean_sd(BD_NOCD, "SCA2_CODE", "DURATION", "Duration BD-NoCD")
wilcox.test(DURATION ~ SCA2_CODE, data = BD_NOCD)

# SCH - SCA2
mean_sd(SCH, "SCA2_CODE", "DURATION", "Duration SCH")
kruskal.test(DURATION ~ SCA2_CODE, data = SCH)

mean_sd(SCH_CD, "SCA2_CODE", "DURATION", "Duration SCH-CD")
kruskal.test(DURATION ~ SCA2_CODE, data = SCH_CD)

mean_sd(SCH_NOCD, "SCA2_CODE", "DURATION", "Duration SCH-NO-CD")
kruskal.test(DURATION ~ SCA2_CODE, data = SCH_NOCD)

# SURVIVAL CURVES ----

# BD
surv_object <- Surv(time = BD$DURATION, event = rep(1, length(BD$DURATION)))
cox_model <- coxph(surv_object ~ SEX + SMOKER + SCA2_CODE + HTT_CODE, data = BD)
summary(cox_model)

survdiff(surv_object ~ COFFEE, data = BD, rho = 0)
survdiff(surv_object ~ SEX, data = BD, rho = 0)
survdiff(surv_object ~ SMOKER, data = BD, rho = 0)
survdiff(surv_object ~ SCA2_CODE, data = BD, rho = 0)
cox_model <- coxph(surv_object ~ SCA2_CODE, data = BD)
summary(cox_model)
survdiff(surv_object ~ HTT_CODE, data = BD, rho = 0)
survdiff(surv_object ~ SCA1_CODE, data = BD, rho = 0)

surv_object <- Surv(time = BD_NOCD$DURATION, event = rep(1, length(BD_NOCD$DURATION)))
survdiff(surv_object ~ SCA2_CODE, data = BD_NOCD, rho = 0)
survdiff(surv_object ~ HTT_CODE, data = BD_NOCD, rho = 0)
survdiff(surv_object ~ SCA1_CODE, data = BD_NOCD, rho = 0)

surv_object <- Surv(time = BD_CD$DURATION, event = rep(1, length(BD_CD$DURATION)))
survdiff(surv_object ~ SCA2_CODE, data = BD_CD, rho = 0)
cox_model <- coxph(surv_object ~ SCA2_CODE, data = BD_CD)
summary(cox_model)
survdiff(surv_object ~ HTT_CODE, data = BD_CD, rho = 0)
survdiff(surv_object ~ SCA1_CODE, data = BD_CD, rho = 0)

surv_object <- Surv(time = BD$DURATION, event = rep(1, length(BD$DURATION)))
gen.km <- survfit(surv_object ~ SCA2_CODE, data = BD, type = "kaplan-meier", error = "tsiatis", conf.type = "log-log", conf.int = 0.95)
plot_km <- ggsurvplot(
  fit = gen.km,
  data = BD,
  conf.int = TRUE,
  pval = TRUE,
  risk.table = TRUE,
  title = "ATXN2 Kaplan-Meier Curve in BD",
  xlab = "Time (years)",
  ylab = "Survival probability",
  legend.title = "ATXN2 Genotype",
  legend.labs = c("Normal", "IAs")
)
print(plot_km)

# SCH
surv_object <- Surv(time = SCH$DURATION, event = rep(1, length(SCH$DURATION)))
cox_model <- coxph(surv_object ~ SEX + SMOKER + SCA2_CODE + HTT_CODE, data = SCH)
summary(cox_model)

survdiff(surv_object ~ SEX, data = SCH, rho = 0)
survdiff(surv_object ~ SMOKER, data = SCH, rho = 0)
survdiff(surv_object ~ SCA2_CODE, data = SCH, rho = 0)
survdiff(surv_object ~ HTT_CODE, data = SCH, rho = 0)
survdiff(surv_object ~ SCA1_CODE, data = SCH, rho = 0)

surv_object <- Surv(time = SCH_NOCD$DURATION, event = rep(1, length(SCH_NOCD$DURATION)))
survdiff(surv_object ~ SCA2_CODE, data = SCH_NOCD, rho = 0)
survdiff(surv_object ~ HTT_CODE, data = SCH_NOCD, rho = 0)
survdiff(surv_object ~ SCA1_CODE, data = SCH_NOCD, rho = 0)

surv_object <- Surv(time = SCH_CD$DURATION, event = rep(1, length(SCH_CD$DURATION)))
survdiff(surv_object ~ SCA2_CODE, data = SCH_CD, rho = 0)
survdiff(surv_object ~ HTT_CODE, data = SCH_CD, rho = 0)
survdiff(surv_object ~ SCA1_CODE, data = SCH_CD, rho = 0)
```

## 07_Enrichr-KG

This script loads and visualizes an interaction network of miRNAs and their target genes, integrating enrichment results from multiple sources (e.g., GO, KEGG, DisGeNET). It provides both an overview of the full network and a filtered subnetwork focusing on *HTT*, *ATXN1*, and *ATXN2* and their immediate neighbors. Nodes are annotated and colored by source, and the network is visualized using both static and interactive layouts with `visNetwork`.

```{r}
nodes <- read.delim(file.choose())   # Select your nodes file
edges <- read.delim(file.choose())   # Select your edges file

head(nodes)
head(edges)

# Initial visualization with visNetwork ----

# Prepare nodes for visNetwork
nodes_vis <- nodes[, c("id", "label", "kind", "color")]
nodes_vis$color <- ifelse(is.na(nodes_vis$color), "gray", nodes_vis$color)  # default to gray if missing

# Prepare edges for visNetwork
edges_vis <- edges
colnames(edges_vis)[colnames(edges_vis) == "source"] <- "from"
colnames(edges_vis)[colnames(edges_vis) == "target"] <- "to"
edges_vis$title <- edges_vis$relation  # shows relation on hover

# Plot interactive network
visNetwork(nodes_vis, edges_vis) %>%
  visPhysics(
    solver = "forceAtlas2Based",
    forceAtlas2Based = list(gravitationalConstant = -150),
    stabilization = list(enabled = TRUE, iterations = 1000),
    minVelocity = 0.1
  ) %>%
  visNodes(   
    color = list(border = "black", background = nodes_vis$color, highlight = "yellow"),
    font = list(size = 16, face = "bold", vadjust = 0),
    scaling = list(min = 10, max = 30)
  ) %>%
  visEdges(smooth = FALSE) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123)

# Highlight target genes (HTT, ATXN1, ATXN2) ----

# Step 1: define target genes and color for each
target_genes <- c("HTT", "ATXN1", "ATXN2")
target_colors <- c("HTT" = "red", "ATXN1" = "blue", "ATXN2" = "green")

# Step 2: get IDs of the target nodes
target_node_ids <- nodes_vis$id[nodes_vis$label %in% target_genes]

# Step 3: initialize edge and node styles
edges_vis$color <- "gray"
edges_vis$width <- 1
edges_vis$dashes <- FALSE
nodes_vis$color_current <- nodes_vis$color

# Step 4: update color and width for each target gene
for (gene in target_genes) {
  gene_id <- nodes_vis$id[nodes_vis$label == gene]
  
  # Update node color
  nodes_vis$color_current[nodes_vis$id == gene_id] <- target_colors[gene]
  
  # Update edges connected to this gene
  idx_edges <- which(edges_vis$from == gene_id | edges_vis$to == gene_id)
  edges_vis$color[idx_edges] <- target_colors[gene]
  edges_vis$width[idx_edges] <- 3
}

# Step 5: visualize with updated styles
visNetwork(nodes_vis, edges_vis) %>%
  visPhysics(
    solver = "forceAtlas2Based",
    forceAtlas2Based = list(gravitationalConstant = -150),
    stabilization = TRUE
  ) %>%
  visNodes(
    color = list(border = "black", background = nodes_vis$color_current),
    font = list(size = 16, face = "bold", vadjust = 0, align = "center"),
    scaling = list(min = 10, max = 40)
  ) %>%
  visEdges(smooth = FALSE) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123)


# Filter network to target genes and neighbors ----

# Rename columns to match igraph requirements
colnames(edges)[1:2] <- c("from", "to")
colnames(nodes)[1] <- "id"

# Create full graph
graph <- graph_from_data_frame(d = edges, vertices = nodes, directed = TRUE)

# Assign labels
V(graph)$label <- nodes$label[match(V(graph)$name, nodes$id)]

# Select target genes
target_nodes <- V(graph)[V(graph)$label %in% target_genes]

# Get 1st-degree neighbors (ego network)
neighbors <- ego(graph, order = 1, nodes = target_nodes, mode = "all")
included_nodes <- unique(unlist(neighbors))
subgraph <- induced_subgraph(graph, vids = included_nodes)

# Assign labels in the subgraph
V(subgraph)$label <- V(graph)$label[match(V(subgraph)$name, V(graph)$name)]


# Assign colors by node type ----

# Define color palette
node_types <- unique(V(subgraph)$kind)
palette_colors <- c(
  "DisGeNET"                            = "#D2DB7D",
  "GO Biological Process 2021"          = "#FFBAFB",
  "Gene"                                = "#C5E1A5",
  "Human Phenotype Ontology"            = "#B6D7FF",
  "KEGG 2021 Human"                     = "#E9E7F0",
  "MGI Mammalian Phenotype Level 4 2021"= "#FF9600"
)

# Prepare subgraph nodes
sub_nodes <- data.frame(
  id = V(subgraph)$name,
  label = V(subgraph)$label,
  kind = V(subgraph)$kind,
  color = palette_colors[V(subgraph)$kind],
  stringsAsFactors = FALSE
)

# Prepare subgraph edges
sub_edges <- igraph::as_data_frame(subgraph, what = "edges")

# Add legend for node types ----

legend_nodes <- data.frame(
  label = names(palette_colors),
  shape = "dot",
  color = unname(palette_colors),
  size = 15,
  stringsAsFactors = FALSE
)

# Simplify legend labels
legend_nodes$label <- c("DisGeNET", "GO BP", "Gene", "HP", "KEGG", "MP")


# Final visualization of subgraph ----

visNetwork(sub_nodes, sub_edges) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visNodes(
    color = list(border = "black", background = sub_nodes$color, highlight = "yellow"),
    borderWidth = 1,
    font = list(face = "bold")
  ) %>%
  visLayout(randomSeed = 123) %>%
  visPhysics(enabled = FALSE) %>%
  visLegend(
    useGroups = FALSE,
    addNodes = legend_nodes,
    position = "left"
  )
```
